The Importance of Initial Conditions in Seasonal Predictions of Antarctic Sea Ice

### 2025-11-27 analysis/paper/access-ice.tex
Renders

Figure~\ref{fig-hindcast-extent} shows mean sea-ice extent of the
ACCESS-S1 and ACCESS-S2 hindcasts (row a) and their differences from
mean sea-ice extent of NOAA/NSIDC CDRV4 (row b). Mean extent at the
first of every month is indicated with circles for the initial
conditions and with triangles for the longest lead time possible for
each model (between 274 and 277 days for ACCESS-S2 and between 213 and
216 days for ACCESS-S1). At this long lead time, information of the
initial conditions is essentially lost and the forecast reverts close to
each model's preferred equilibrium state.

ACCESS-S2 initial conditions (circles in Fig.~\ref{fig-hindcast-extent}
column 2) show an overall negative bias, especially in the late
summer-early autumn, while ACCESS-S1 initial conditions (circles in
Fig.~\ref{fig-hindcast-extent} column 1) are very close to observations,
as expected from the assimilation of sea-ice observations to produce the
initial conditions of ACCESS-S1. Both systems' equilibrium states
(triangles) show negative biases of sea-ice extent, particularly in the
growth phase of late-autumn and winter months. This is due primarily to
the melt season being longer than in observations and with faster melt
between January and March and the growing seasons being shorter with
slower growth during March and April. This is then followed by faster
growth between May and July (Figure~\ref{fig-mean-growth}). Many sea-ice
models exhibit this systematic underestimation during the sea-ice
minimum and early freezing season \citep{massonnet2023}, which could
indicate problems in the representation of thermodynamics in the model
\citep{zampieri2019}. It is also not surprising that both forecasting
systems converge to a similar equilibrium state because they share the
same model formulation.

@@

To further understand the bias in ACCESS-S2, Figure~\ref{fig-bias-1}
shows spatial patterns of the differences of monthly mean sea-ice
concentrations between NOAA/NSIDC CDRV4 and ACCESS-S2 hindcasts at the
shortest monthly lead time. From October to May, the model
underestimates sea-ice concentrations in most regions except for the
inner Weddell Sea in April and May, where sea-ice concentrations
saturate to 1 both in the observations and forecasts. In winter, the
differences are limited to a narrow band around the sea-ice edge with
slight positive biases in the African sector of East Antarctica and
negative biases around the Indian Ocean sector which partially
compensate, resulting in the near-zero extent bias seen in those months
(Figure~\ref{fig-hindcast-extent}).

ACCESS-S1 has a comparatively smaller overall bias
(Figure~\ref{fig-bias-2}). The largest values are found between April
and June, when the faster growth results in large positive bias along
the sea-ice edge, and in January, when the faster melt leads to large
negative bias in the Weddell and Amundsen Seas.

@@

Figure~\ref{fig-extent-anom} shows monthly sea-ice extent anomalies
forecasted at selected lead times. Compared with ACCESS-S1, ACCESS-S2
anomaly forecasts are relatively poor (large RMSE) even for the first
month (lead time 0), whereas ACCESS-S1 forecasts stay relatively skilful
even at a lead time of three months. ACCESS-S2 shows much larger
interannual variability than observations, with dramatic lows between
1995 and 2007, and highs between 2007 and 2015.

Unexpectedly, for ACCESS-S2, RMSE improves with lead time, even though
the correlation degrades with lead time. This is puzzling behaviour that
goes contrary to what is usually seen in prediction models. The
explanation seems to be the mentioned increased interannual variability.
Figure~\ref{fig-extent-sd} shows the interannual standard deviation of
monthly sea-ice extent of the forecasts as a function of lead time
compared with observations. ACCESS-S1 standard deviation lies within the
observed standard deviation regardless of lead time, while ACCESS-S2
standard deviation is more than twice that of observations at zero lead
time and only approaches the observed value at nine month lead time for
most months.

@@

To assess ACCESS-S2 forecasts in more detail, we compute error measures
for all hindcasts started on the 1st of every month.
Figure~\ref{fig-rmse} shows the mean RMSE of sea-ice concentration
anomalies for ACCESS-S1 and ACCESS-S2 hindcasts compared against
persistence and climatological forecasts used as a benchmark. Due to
errors in the initial conditions, it is expected that persistence
forecasts would be better than the model forecasts at very short lead
times, but that the persistence forecast errors would grow faster and
may eventually surpass the model forecast errors. The black line shows
that the persistence forecast error indeed grows rapidly and reaches its
maximum in about 30 days for most months except for February, when it
grows much slower. The ACCESS-S1 forecast errors grow slower than
persistence forecast errors and remain lower after less than 10 days on
average. The ACCESS-S2 forecast error starts high in all months and is
lower than the persistence forecast error after more than 15 days in
most months except for forecast initialised in February, when it takes
80 days.

@@

Figure~\ref{fig-lead-time-window} summarises the lead time window in
which each hindcast is better than both the persistence forecast and the
climatological forecast as a function of forecast month. ACCESS-S1
forecasts have a wider lead time window in the summer than the other
seasons and is not better than both benchmarks at forecasting June
sea-ice concentration anomalies. Forecasts initialised in May and June
are particularly poor, and July cannot be forecasted better than the
benchmarks. This is consistent with the mid-winter loss of
predictability observed by \citet{libera2022}, who attributed it to deep
warm water entraining into the mixed layer.

@@

For ACCESS-S1 forecasts (Figure~\ref{fig-rmse_lon-1}), skill tends to be
lower off the coast of Eastern Antarctica even at short lead times; for
instance, the skill score for forecasts initialised in May and June are
negative between 0° and 120°E even at almost zero lead time. This
mirrors \citet{libera2022} findings of a ``winter predictability
barrier'', although they focus on the Weddell Sea and here we show that
the effect seems to be stronger more to the east. In West Antarctica
there is a hint of easterly-propagating skill in forecasts initialised
in February and March. This is consistent with \citet{holland2013}
findings that memory of sea-ice anomalies are stored in ocean heat
content anomalies that are transported east by the Antarctic Circumpolar
Current.

ACCESS-S2 forecasts (Figure~\ref{fig-rmse_lon-2}) also have lower skill
over East Antarctica. From July to December even though the
pan-Antarctic average skill is negative at all lead times
(Fig.~\ref{fig-lead-time-window}), it is positive for up to a month in
West Antarctica. Since oceanic and atmospheric forcing is the only
source of information, this suggests that sea-ice in this region is
particularly sensitive to oceanic and atmospheric forcing and suggests a
role of the Pacific-South American mode and the Amundsen Sea Low to
shape sea-ice concentration anomalies. The fact that this is evident in
the months in which El Niño--Southern Oscillation teleconnections are
more important for atmospheric circulation also suggests the influence
of tropical Pacific variability. February and March are the only two
months that can be forecasted with marginally positive skill in large
regions.

@@

Finally, Figure~\ref{fig-rmse_lon-3} shows the difference in skill
between ACCESS-S1 and ACCESS-S2. Large differences in skill indicate
areas and months that are most affected by the data assimilation present
in ACCESS-S1. Between January and March, which are the months in which
ACCESS-S1 is the most skilful (Figure~\ref{fig-lead-time-window}), most
of the improvement compared with ACCESS-S2 is present in the Ross and
Weddell Sea. In April and May, the improvement seems more homogeneous.

@@

In Figure~\ref{fig-rmse} the mean error was shown.
Figure~\ref{fig-initial-spread} column 1 shows the mean standard
deviation of errors among ensemble members at various lead times. At one
day lead time (Fig.~\ref{fig-initial-spread} a.1) ACCESS-S2 has a
slightly larger spread than ACCESS-S1 due to the way that ensemble
members are generated. ACCESS-S1 ensemble members are generated by
adding random field perturbations to the atmosphere only, which then are
transferred to the other components via the coupled simulation
\citep{hudson2017}. With this scheme, ensemble members are all but
guaranteed to be underdispersed in the ocean and sea-ice components. The
time-lag ensemble used for ACCESS-S2 ensures greater spread. This
difference is gone after about just two days, and both systems have a
comparable spread in ensemble member error afterwards
(Fig.~\ref{fig-initial-spread} b1 and c1).

>>>>

Figure~\ref{fig-hindcast-extent} (Figure~\ref{fig-hindcast-extent-osi}
for OSI) shows mean sea-ice extent of the ACCESS-S1 and ACCESS-S2
hindcasts (row a) and their differences from mean sea-ice extent of
NOAA/NSIDC CDRV4 (row b). Mean extent at the first of every month is
indicated with circles for the initial conditions and with triangles for
the longest lead time possible for each model (between 274 and 277 days
for ACCESS-S2 and between 213 and 216 days for ACCESS-S1). At this long
lead time, information of the initial conditions is essentially lost and
the forecast reverts close to each model's preferred equilibrium state.

ACCESS-S2 initial conditions (circles in Fig.~\ref{fig-hindcast-extent}
column 2) show an overall negative bias, especially in the late
summer-early autumn, while ACCESS-S1 initial conditions (circles in
Fig.~\ref{fig-hindcast-extent} column 1) are very close to observations,
as expected from the assimilation of sea-ice observations to produce the
initial conditions of ACCESS-S1. Both systems' equilibrium states
(triangles) show negative biases of sea-ice extent, particularly in the
growth phase of late-autumn and winter months. This is due primarily to
the melt season being longer than in observations and with faster melt
between January and March and the growing seasons being shorter with
slower growth during March and April. This is then followed by faster
growth between May and July (Figure~\ref{fig-mean-growth} and
Figure~\ref{fig-mean-growth-osi}). Many sea-ice models exhibit this
systematic underestimation during the sea-ice minimum and early freezing
season \citep{massonnet2023}, which could indicate problems in the
representation of thermodynamics in the model \citep{zampieri2019}. It
is also not surprising that both forecasting systems converge to a
similar equilibrium state because they share the same model formulation.

@@

To further understand the bias in ACCESS-S2, Figure~\ref{fig-bias-1}
(Figure~\ref{fig-bias-1-osi}) shows spatial patterns of the differences
of monthly mean sea-ice concentrations between NOAA/NSIDC CDRV4 and
ACCESS-S2 hindcasts at the shortest monthly lead time. From October to
May, the model underestimates sea-ice concentrations in most regions
except for the inner Weddell Sea in April and May, where sea-ice
concentrations saturate to 1 both in the observations and forecasts. In
winter, the differences are limited to a narrow band around the sea-ice
edge with slight positive biases in the African sector of East
Antarctica and negative biases around the Indian Ocean sector which
partially compensate, resulting in the near-zero extent bias seen in
those months (Figure~\ref{fig-hindcast-extent}).

ACCESS-S1 has a comparatively smaller overall bias
(Figure~\ref{fig-bias-2} and Figure~\ref{fig-bias-2-osi}). The largest
values are found between April and June, when the faster growth results
in large positive bias along the sea-ice edge, and in January, when the
faster melt leads to large negative bias in the Weddell and Amundsen
Seas.

@@

Figure~\ref{fig-extent-anom} (Figure~\ref{fig-extent-anom-osi}) shows
monthly sea-ice extent anomalies forecasted at selected lead times.
Compared with ACCESS-S1, ACCESS-S2 anomaly forecasts are relatively poor
(large RMSE) even for the first month (lead time 0), whereas ACCESS-S1
forecasts stay relatively skilful even at a lead time of three months.
ACCESS-S2 shows much larger interannual variability than observations,
with dramatic lows between 1995 and 2007, and highs between 2007 and
2015.

Unexpectedly, for ACCESS-S2, RMSE improves with lead time, even though
the correlation degrades with lead time. This is puzzling behaviour that
goes contrary to what is usually seen in prediction models. The
explanation seems to be the mentioned increased interannual variability.
Figure~\ref{fig-extent-sd} (Figure~\ref{fig-extent-sd-osi}) shows the
interannual standard deviation of monthly sea-ice extent of the
forecasts as a function of lead time compared with observations.
ACCESS-S1 standard deviation lies within the observed standard deviation
regardless of lead time, while ACCESS-S2 standard deviation is more than
twice that of observations at zero lead time and only approaches the
observed value at nine month lead time for most months.

@@

To assess ACCESS-S2 forecasts in more detail, we compute error measures
for all hindcasts started on the 1st of every month.
Figure~\ref{fig-rmse} (Figure~\ref{fig-rmse-osi}) shows the mean RMSE of
sea-ice concentration anomalies for ACCESS-S1 and ACCESS-S2 hindcasts
compared against persistence and climatological forecasts used as a
benchmark. Due to errors in the initial conditions, it is expected that
persistence forecasts would be better than the model forecasts at very
short lead times, but that the persistence forecast errors would grow
faster and may eventually surpass the model forecast errors. The black
line shows that the persistence forecast error indeed grows rapidly and
reaches its maximum in about 30 days for most months except for
February, when it grows much slower. The ACCESS-S1 forecast errors grow
slower than persistence forecast errors and remain lower after less than
10 days on average. The ACCESS-S2 forecast error starts high in all
months and is lower than the persistence forecast error after more than
15 days in most months except for forecast initialised in February, when
it takes 80 days.

@@

Figure~\ref{fig-lead-time-window}
(Figure~\ref{fig-lead-time-window-osi}) summarises the lead time window
in which each hindcast is better than both the persistence forecast and
the climatological forecast as a function of forecast month. ACCESS-S1
forecasts have a wider lead time window in the summer than the other
seasons and is not better than both benchmarks at forecasting June
sea-ice concentration anomalies. Forecasts initialised in May and June
are particularly poor, and July cannot be forecasted better than the
benchmarks. This is consistent with the mid-winter loss of
predictability observed by \citet{libera2022}, who attributed it to deep
warm water entraining into the mixed layer.

@@

For ACCESS-S1 forecasts (Figure~\ref{fig-rmse_lon-1} and
Figure~\ref{fig-rmse_lon-1-osi}), skill tends to be lower off the coast
of Eastern Antarctica even at short lead times; for instance, the skill
score for forecasts initialised in May and June are negative between 0°
and 120°E even at almost zero lead time. This mirrors \citet{libera2022}
findings of a ``winter predictability barrier'', although they focus on
the Weddell Sea and here we show that the effect seems to be stronger
more to the east. In West Antarctica there is a hint of
easterly-propagating skill in forecasts initialised in February and
March. This is consistent with \citet{holland2013} findings that memory
of sea-ice anomalies are stored in ocean heat content anomalies that are
transported east by the Antarctic Circumpolar Current.

ACCESS-S2 forecasts (Figure~\ref{fig-rmse_lon-2} and
Figure~\ref{fig-rmse_lon-2-osi}) also have lower skill over East
Antarctica. From July to December even though the pan-Antarctic average
skill is negative at all lead times (Fig.~\ref{fig-lead-time-window}),
it is positive for up to a month in West Antarctica. Since oceanic and
atmospheric forcing is the only source of information, this suggests
that sea-ice in this region is particularly sensitive to oceanic and
atmospheric forcing and suggests a role of the Pacific-South American
mode and the Amundsen Sea Low to shape sea-ice concentration anomalies.
The fact that this is evident in the months in which El Niño--Southern
Oscillation teleconnections are more important for atmospheric
circulation also suggests the influence of tropical Pacific variability.
February and March are the only two months that can be forecasted with
marginally positive skill in large regions.

@@

Finally, Figure~\ref{fig-rmse_lon-3} (Figure~\ref{fig-rmse_lon-3-osi})
shows the difference in skill between ACCESS-S1 and ACCESS-S2. Large
differences in skill indicate areas and months that are most affected by
the data assimilation present in ACCESS-S1. Between January and March,
which are the months in which ACCESS-S1 is the most skilful
(Figure~\ref{fig-lead-time-window}), most of the improvement compared
with ACCESS-S2 is present in the Ross and Weddell Sea. In April and May,
the improvement seems more homogeneous.

@@

In Figure~\ref{fig-rmse} the mean error was shown.
Figure~\ref{fig-initial-spread} (Figure~\ref{fig-initial-spread-osi})
column 1 shows the mean standard deviation of errors among ensemble
members at various lead times. At one day lead time
(Fig.~\ref{fig-initial-spread} a.1) ACCESS-S2 has a slightly larger
spread than ACCESS-S1 due to the way that ensemble members are
generated. ACCESS-S1 ensemble members are generated by adding random
field perturbations to the atmosphere only, which then are transferred
to the other components via the coupled simulation \citep{hudson2017}.
With this scheme, ensemble members are all but guaranteed to be
underdispersed in the ocean and sea-ice components. The time-lag
ensemble used for ACCESS-S2 ensures greater spread. This difference is
gone after about just two days, and both systems have a comparable
spread in ensemble member error afterwards
(Fig.~\ref{fig-initial-spread} b1 and c1).
